---
title: "CMS-HCC on MEPS"
author: C. Seth Lester, ASA, MAAA ([cslester@gmail.com](mailto:cslester@gmail.com)) <br>
date: 29 September 2024 (nonfinal version)
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Custom color pallette
cp <- rev(c('#d73027','#fc8d59','#fee090','#ffffbf','#e0f3f8','#91bfdb','#4575b4'))

```

# Project Overview

The code in this repository applies the CMS-HCC risk adjustment model to data from the Medical Expenditure Panel Survey (MEPS) Public Use Files (PUFs) for the years 2016 - 2022. 

## What is Risk Adjustment in Healthcare?

Healthcare risk adjustment models, such as the CMS-HCC risk adjustment model, are designed to take administrative healthcare claims data for the current year as inputs and produce individual-level risk scores as outputs. The risk score output for each individual processed by the CMS-HCC model is designed to be highly correlated with that individual's total costs to the Medicare program in the subsequent year.

## What is the CMS-HCC model used for?

The CMS-HCC model is used to adjust payments to Medicare Advantage plans based on the health status and demographic characteristics of enrollees, ensuring that plans are compensated fairly for taking on higher-risk patients. Without such a risk adjustment-based reimbursement model, Medicare Advantage plans would be incentivized against enrolling high-risk patients or those members more generally in need of costly healthcare services, such as members with pre-existing (but manageable) chronic conditions like diabetes, high blood pressure, or asthma.

On the contrary, one could argue that this reimbursement model over-incentivizes Medicare Advantage plans to take on high-risk patients [(or potentially report diagnoses in patients that may bend the truth, so that they appear to be high-risk)](https://www.wsj.com/health/healthcare/medicare-health-insurance-diagnosis-payments-b4d99a5d), but this is a contentious discussion, and this repository is not the place for it.

## What else can the CMS-HCC model be used for?

Beyond adjusting payments to Medicare Advantage plans, the CMS-HCC model is often used for other analytical purposes:

- **Projecting Future Costs**: Organizations frequently apply the CMS-HCC model to claims data to project future costs, even when not directly tied to reimbursement. For example, a Medicare Advantage plan might use these prospective risk scores to project costs for their enrollee population in the coming year. *Vignette coming soon!*

- **Controlling for Clinical Risk in Observational Studies**: The CMS-HCC risk adjuster is also used in observational studies to control for or quantify clinical risk differences between two populations, such as treatment and control groups. Since the CMS-HCC model is trained prospectively on costs incurred in the following year, it is particularly useful for calculating average treatment effects in terms of ROI or dollars saved by some program, intervention, or social disparity experienced by the treatment population and not experienced by the control population.  *Vignette coming soon!*

## Why MEPS?

The Medical Expenditure Panel Survey (MEPS) is [widely used in health outcomes and effectiveness studies](https://meps.ahrq.gov/mepsweb/data_stats/publications.jsp), making it a valuable resource for healthcare research. MEPS collects detailed information on healthcare utilization, expenditures, insurance coverage, and the health status of individuals, families, and their providers across the U.S., as well as socio-demographic information, family size and structure data, and individual responses to screening instruments like the [Veterans RAND 12 Item Health Survey](https://www.bu.edu/sph/research/centers-and-groups/vr-36-vr-12-and-vr-6d/about-the-vr-36-vr-12-and-vr-6d/) (VR-12), [Patient Health Questionnaire-2](https://www.hiv.uw.edu/page/mental-health-screening/phq-2) (PHQ-2), or other ad-hoc screening instruments (including a comprehesive battery of [screening questions evaluating SDoH factors](https://meps.ahrq.gov/survey_comp/SHE_Sources.pdf) administered to panels 23 - 26). 

While MEPS data comes with the usual litany of limitations inherent in a large-scale longitudinal survey — such as missing data, imputed cost values in the absence of responses, non-response bias, and other potential biases driven by respondent errors, etc., it remains one of the few large-scale public data resources that contains a subset of data with the same "look and feel" of administrative claims data. 

## MEPS + the CMS-HCC Model = a perfect-*ish* pairing!

With a little analytic finagling, MEPS datasets have all the basic inputs required by the CMS-HCC risk adjustment model! This includes each respondent's exact age (down to the month and year of birth, top-coded at age 85), gender, and health conditions expressed in the same ICD-10-CM ontology used to express diagnosis codes in administrative claims data.

Additionally, MEPS is a longitudinal survey. This means that the survey instrument is applied to an [impaneled group of respondents and followed over multiple years](https://www.meps.ahrq.gov/survey_comp/hc_data_collection.jsp). This is essential because the CMS-HCC risk adjustment model is a prospective model - its intended use is to produce a risk score correlated with individual total cost to Medicare based on medical diagnosis information drawn from the previous year. 

Without MEPS' longitudinal design, it would be impossible to measure the predictive performance of the CMS-HCC risk adjuster model since this requires, at least, a year of diagnosis information and then information about Medicare expenditures in the subsequent year. MEPS has this information!

## What's in this repository?

This code repository contains all the R and Python code responsible for the "analytic finagling" required to process MEPS respondent data using version 24 and 28 of the CMS-HCC model. The CMS-HCC model outputs available are the CMS-HCC raw risk scores assigned to each individual respondent for the NE (New Enrollee) and the Community Non-Dual Aged (CNA) models.

Both the NE and CNA models for CMS-HCC versions 24 and 28 are processed on all MEPS respondents with at least one 1 month of coverage under Medicare, as defined by the variables *MCRmmyy* present in each year's Full Year Consolidated (FYC) public use data file, where *mm* represents a two-letter month code, and *yy* represents a two digit year. 

More information about these variables (and other respondent-level data available in MEPS) can be accessed using the [MEPS-HC Data Tools Variable Explorer](https://datatools.ahrq.gov/meps-hc?_gl=1*32wp2d*_ga*MTY5NDQ1MTMyLjE3MTIwNjgwODU.*_ga_1NPT56LE7J*MTcyNzU3ODg4NC4zMy4xLjE3Mjc1ODEzNzguMC4wLjA.#varexpLabel).

The CMS-HCC model implementation used to process respondents in this repository is a unique Python implementation of the CMS-HCC model, [hccpy](https://github.com/yubin-park/hccpy). Outputs generated by this model implementation has not been tested against the official [SAS implemenation of the CMS-HCC model](https://www.nber.org/research/data/risk-adjustment).

## If I don't care about the process, can I just download the risk scores for each member to use for my own MEPS analysis project?

Sure, here's a zipfile of all the output generated from this repo, along with some very light documentation. Enjoy!

# Assessing CMS-HCC Model Performance on MEPS Data

To ensure this project isn't a completely futile exercise, I have compiled several visualizations that describe the CMS-HCC models' predictive performance measurements on MEPS data, confirming acceptable performance through metrics like R-squared and Actual-to-Expected (A/E) ratios, which underscore the primary goal of a risk adjuster model - predicting future healthcare costs.

While we can generally expect that risk adjuster models might behave differently (and, sometimes, *poorly*) when exposed to data generated by populations that were not used in their initial training and development, this project would be quite a silly exercise if the predictive performance mesaurements of the CMS-HCC risk adjusters on the MEPS input data were not at least somewhat comparable to the predictive performance measures of the CMS-HCC model published in the most recent (2021) [report to Congress required by Section 17006(f)(2)(A)(ii) of the 21st Century Cures Act](https://www.cms.gov/files/document/report-congress-risk-adjustment-medicare-advantage-december-2021.pdf).

## Measuring Model Performance - Determining the Applicable MEPS Target Population

In this section, I discuss how we can apply filters to MEPS variables in order to determine the subset of MEPS respondents for which it is appropriate to process using the CMS-HCC NE (new enrollee) and CNA (community non-dual aged) models.

The MEPS PUFs contain data on a wide variety of respondents with varying types of health plan coverage, and also includes uninsured respondents with no health plan coverage at all. Respondents in the MEPS PUFs are each assigned a weight as part of a probability sampling frame, since respondents are not uniformly representative of the MEPS target population, which is all non-institutionalized civilians in the United States. So, it is important to remember that while we are working directly with real respondent data in this project, our unweighted sample population is *arbitrary* and not necessarily representative of any target population.

1. **Members in the target population should have two years of coverage in MEPS and Medicare coverage for at least one month in the second year.** For example, for 2016 MEPS data, we find all respondents who are impaneled in both 2016 and 2017 MEPS data, and then filter these respondents on those with at least one month of Medicare coverage in 2017. This process described for the year 2016 is then repeated for years 2017 - 2021. Medicare coverage is defined in the MEPS data as a value of 1 for the variable series *MCRmmyy*, where *mm* is a two-letter representation of the month, and *yy* is a two-digit representation of the year.

2. **The target population should not contain dual-eligible respondents.** Since we are measuring the performance of the CMS-HCC CNA model exclusively, our applicable population should have not contain individuals with a nonzero number of months of Medicaid coverage in the subsequent year. As in (1) above, we filter out this population by removing all respondents with a value of 1 for any of the 12 monthly Medicaid coverage variables in the payment year.

3. **The target population should should be age-filtered to remove as many respondents as possible who may have originally been entitled to Medicare benefits due to disability status.** Since we are interested in measuring the performance of the CMS-HCC risk adjuster models using the CNA model, our applicable population should not contain respondents who are or originally were covered by Medicare due to disability. Since there is no high-fidelity indicator of disability status in the MEPS data, we resort to filtering on members whose exact age is 65 or greater as of 1/1/YY, where YY is the payment year. This isn't perfect as it will include benes who originally qualified for Medicare due to disability and have stayed on Medicare past their 65th birthday, but it's the best we can do.
<br><br>**The target population should contain only individuals under age 85.** For privacy reasons, respondent ages in MEPS PUFs are top-coded at age 85. This means that any respondent *over* the age of 85 will show as actually age 85 for any year of data. This is problematic for our project because the CMS-HCC model has age/gender weightings for individuals in 85-89, 90-94, and over 95 age buckets. If we do not remove individuals whose age is top-coded, then they will be assigned risk scores consisting of an incorrect age/gender weighting, which will reduce overall model performance.

4. **The target population should not contain respondents with an ESRD diagnosis.** There is a separate CMS-HCC model intended for use on this population. Individuals with ESRD tend to have much larger expenditures than those without an ESRD diagnosis, and thus it would not be fair to measure the CMS-HCC model performance with these members present in the target population.

The following visualization shows, grouped by MEPS data year, the number of respondents included in each successive wave of filtering described above. Ultimately, for each year, there are around 1,500 - 2,000 respondents who make it through the filtration steps and processed by versions 24 and 28 of the CMS-HCC's new enrollee (NE) and Community Non-Dual Aged (CNA) models.

```{r describe_target_pop_unwt, echo=FALSE, fig.height=8.5, fig.width=11, message=FALSE, warning=FALSE}

library(tidyverse)
library(haven)
library(MEPS)
library(survey)
library(srvyr)

options(survey.lonely.psu='adjust')

# Read in survey weights information
survey_weights <- read_csv(here::here("etc/outputs/survey_weights.csv")) %>% 
  filter(meps_year <= 2021)

# Layer in various levels of filtration
filter_step1 <- read_csv(here::here("etc/outputs/mcr_benes.csv")) %>% 
  mutate(step1_filter = 1)

filter_step2 <- read_csv(here::here("etc/outputs/mcr_benes_no_duals.csv")) %>% 
  mutate(step2_filter = 1)

filter_step3 <- read_csv(here::here("etc/outputs/demographics.csv")) %>% 
  select(DUPERSID, meps_year) %>% 
  mutate(step3_filter = 1)

esrd_individuals <- read_csv(here::here("etc/outputs/esrd_individuals.csv"))
filter_step4 <- filter_step3 %>% 
  anti_join(esrd_individuals) %>% 
  mutate(step4_filter = 1)

survey_weights_filtered <- survey_weights %>% 
  left_join(filter_step1) %>% 
  left_join(filter_step2) %>% 
  left_join(filter_step3) %>% 
  left_join(filter_step4) %>% 
  replace_na(list(step1_filter=0,
                  step2_filter=0,
                  step3_filter=0,
                  step4_filter=0)) 

unweighted_counts.data <- survey_weights_filtered %>% 
  group_by(meps_year) %>% 
  summarize(step0_count = n(),
            step1_count = sum(step1_filter),
            step2_count = sum(step2_filter),
            step3_count = sum(step3_filter),
            step4_count = sum(step4_filter)) %>% 
  pivot_longer(
    cols = -c(meps_year),
    names_to = "step",
    values_to = "count",
    names_pattern = "^step(0|1|2|3|4)_count$"
  ) %>% 
  mutate(step_dsc = case_when(
    step == 0 ~ str_wrap("Total Number of MEPS Respondents", 25),
    step == 1 ~ str_wrap("Step 1: Respondents with 2 years of data and at least one month of Medicare coverage in the 2nd year", 25),
    step == 2 ~ str_wrap("Step 2: Respondents from Step 1 not covered by Medicaid in the 2nd year", 25),
    step == 3 ~ str_wrap("Step 3: Respondents from Step 2 ages 65 - 84", 25),
    step == 4 ~ str_wrap("Step 4: Respondents from Step 3 not diagnosed with ESRD", 25))) %>% 
  arrange(step) %>% 
  mutate(step_dsc = fct_inorder(step_dsc)) %>% 
  arrange(meps_year) %>% 
  mutate(meps_year = fct_inorder(as.character(meps_year))) 

unweighted_counts.plot <- unweighted_counts.data %>% 
  filter(step != 0) %>% 
  ggplot(aes(x = meps_year, y = count, fill = step_dsc)) +
  geom_bar(stat = "identity", color="gray20", 
           position=position_dodge2(width=.9)) +
  geom_text(aes(label = scales::label_comma(accuracy=1)(count)),
             position = position_dodge2(width=.9),
             size = 3, vjust = -.8) +
  scale_fill_manual(values = cp) + 
  scale_y_continuous(labels = scales::label_comma(accuracy=1),
                     breaks = c(0, 6000)) +
  theme_bw() +
  labs(
    title = "MEPS Respondent Filtering Stages for CMS-HCC Risk Scoring",
    subtitle = "Unweighted Sample Sizes",
    x = "MEPS Data Year",
    y = "Count of Respondents"
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  )

unweighted_counts.plot

```

It is important to remember that respondents in MEPS are not uniformly weighted. The following visualization gives a view of *weighted* sample sizes risk scored by this project in each year. In other words, the roughly 1,500 - 2,000 respondents scored in each year are intended to be representative of roughly 20 - 30 million members of the US civilian non-institutionalized population in each year.

```{r describe_target_pop_wt, echo=FALSE, fig.height=8.5, fig.width=11, message=FALSE, warning=FALSE}
# Define helper function to create survey design objects from datasets
svydsn_helper <- function(dataset) {
  dataset %>% 
    srvyr::as_survey_design(
      id = VARPSU,
      strata = VARSTR,
      weights = PERWTYYF,
      nest = T
    ) %>% 
    srvyr::summarize(step1_count = srvyr::survey_total(step1_filter, vartype=c("se", "ci"), df=Inf),
                     step2_count = srvyr::survey_total(step2_filter, vartype=c("se", "ci"), df=Inf),
                     step3_count = srvyr::survey_total(step3_filter, vartype=c("se", "ci"), df=Inf),
                     step4_count = srvyr::survey_total(step4_filter, vartype=c("se", "ci"), df=Inf)) %>% 
    select(matches("^step(1|2|3|4)_count$"), matches("^step(1|2|3|4)_count_(low|upp)$"))
}

# Now to do the weighted counts
weighted_counts.data <- survey_weights_filtered %>% 
  group_by(meps_year) %>% 
  nest() %>% 
  mutate(weighted_summary_table = map(data, svydsn_helper)) %>% 
  select(-data) %>% 
  unnest(cols = weighted_summary_table) %>% 
  rename_with(~paste0(., "_est"), .cols = matches("^step(1|2|3|4)_count$")) %>% 
  ungroup() %>% 
  pivot_longer(
    cols = -c(meps_year),
    names_to = c("step", "label"),
    values_to = "value",
    names_pattern = "^step(1|2|3|4)_count_(est|low|upp)$"
  ) %>% 
  pivot_wider(
    names_from = "label",
    values_from = "value",
  ) %>% 
  mutate(step_dsc = case_when(
    step == 1 ~ str_wrap("Step 1: Respondents with 2 years of data and at least one month of Medicare coverage in the 2nd year", 25),
    step == 2 ~ str_wrap("Step 2: Respondents from Step 1 not covered by Medicaid in the 2nd year", 25),
    step == 3 ~ str_wrap("Step 3: Respondents from Step 2 ages 65 - 84", 25),
    step == 4 ~ str_wrap("Step 4: Respondents from Step 3 not diagnosed with ESRD", 25))) %>% 
  arrange(step) %>% 
  mutate(step_dsc = fct_inorder(step_dsc)) %>% 
  arrange(meps_year) %>% 
  mutate(meps_year = fct_inorder(as.character(meps_year))) 

weighted_counts.plot <- weighted_counts.data %>% 
  ggplot(aes(x = meps_year, y = est, fill = step_dsc)) +
  geom_bar(stat = "identity", color="gray20", 
           position=position_dodge2(width=.9)) +
  geom_text(aes(label = scales::label_comma(accuracy=.1, scale=1e-6, suffix=" M")(est)),
             position = position_dodge2(width=.9),
             size = 3, vjust = -6) +
  scale_color_manual(values = c(error = "black"), labels = c("error" = "Error bars denote 95% CI")) +
  geom_errorbar(aes(color = "error", ymin = low, ymax = upp), 
                position = position_dodge(.9), width = .2) +
  scale_fill_manual(values = cp) + 
  scale_y_continuous(labels = scales::label_comma(accuracy=1, scale=1e-6, suffix = " million")) +
  theme_bw() +
  labs(
    title = "MEPS Respondent Filtering Stages for CMS-HCC Risk Scoring",
    subtitle = "Weighted Sample Sizes",
    x = "MEPS Data Year",
    y = "Weighted Sample Sizes"
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  )

weighted_counts.plot

```

## Predictive Performance of New Enrollee Model

The CMS-HCC New Enrollee (NE) model is intended for use on Medicare beneficiaries with at least one month of Medicare coverage in the payment year (where "payment year" is the 12 month period following the 12 month period from which the administrative claims data or demographics information was gathered). This model does not require any medical diagnosis code information because it is designed to produce a prospective risk score using only the member's age and gender to predict future Medicare expenditures. Because this model uses so little information to produce a predictive risk score, it is generally accepted that NE models are not particularly predictive and we should expect a relatively low R-squared value from this model's predictions for all years.

The most recent report to Congress on the CMS-HCC risk adjustment model does not include any R-squared measurements that can be used as a reference comparison against our observations of the predictive performance of the NE model on MEPS data. However, we can look to past inceptions of risk adjuster models - namely the [legacy AAPCC model](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4193587/) used for Medicare risk adjustment prior to the year 2000 - which, similar to the CMS-HCC NE model, used demographic information alone to predict healthcare expenditures. Per the most recent [report to Congress](https://www.cms.gov/files/document/report-congress-risk-adjustment-medicare-advantage-december-2021.pdf), Table 2-1, we might expect something in the ballpark of .77% for an R-squared measuremet on this model for any given year.

### New Enrollee Model - R-Squared for Entire Population, by MEPS Year

Out of the gate, we can easily evaluate the predictive performance of both versions 24 and 28 of the CMS-HCC risk adjuster new enrollee (NE) models on MEPS data. For this exercise, we'll use the square of the correlation between raw risk score output from the NE model for each input year and individual PMPM costs in the subsequent year as the (widely used) definition of R-squared.

```{r rsq_ne, echo=FALSE, fig.height=8.5, fig.width=11, message=FALSE, warning=FALSE}

# Read in cost projections 
cna_pop_pmpm_pred <- read_csv(here::here("etc/outputs/cna_population_pred_costs.csv"))

#### R-Squared Plot for NE Models by MEPS Year ####
# Evaluate r-sq by meps_year for NE. Should focus only on the pop. for which the NE is intended.
rsq_eval_NE.data <- cna_pop_pmpm_pred %>% 
  filter(model == "NE") %>% 
  group_by(meps_year, version) %>% 
  summarize(rsq = cor(cost_pmpm_act, cost_pmpm_pred)^2,
            count = n()) %>% 
  ungroup() %>% 
  mutate(meps_year = meps_year - 1) #Since this is an evaluation of scores, more intuitive to frame in terms of base year
  
rsq_eval_NE.plot <- rsq_eval_NE.data %>% 
  mutate(meps_year = fct_inorder(paste0(meps_year, "\n(n = ", scales::label_comma(accuracy=1)(count), ")"))) %>% 
  mutate(model = paste0("CMS-HCC v", version, " NE")) %>% 
  ggplot(aes(x = meps_year,
             y = rsq,
             fill = model)) +
  geom_bar(stat = "identity", position="dodge2",
           color="gray20") + 
  geom_text(aes(label=scales::label_percent(accuracy=.01)(rsq)),
            position=position_dodge2(width=.9), vjust=-.8, size=3) +
  theme_classic() +
  scale_fill_manual(values = cp) +
  scale_y_continuous(labels = scales::label_percent(accuracy=.1)) +
  labs(title = "Predictive Performance of CMS-HCC New Enrollee Models",
       subtitle = "Using Medical Expenditure Panel Data (MEPS) PUFs Datasets as Inputs",
       y = "R-Squared",
       x = "MEPS Data Input Year",
       fill = "Model")

rsq_eval_NE.plot
```

Unsurprisingly there is very crummy R-squared measured for input year 2019, because expenditure data for 2020 was likely very difficult to accurately survey during the pandemic.

### New Enrollee Model - Predictive Ratios for Entire Population, by Age and Gender

One of the most important uses of a risk adjuster model is to accurately predict costs. We can plot predicted vs. actual costs, grouped by age and gender, in order to evaluate how close (or far off) the predicted PMPM costs for each age band are, across all years evaluated.

```{r a2e_ne_age_gender, echo=FALSE, fig.height=8.5, fig.width=11, message=FALSE, warning=FALSE}

# Read in cost projections 
cna_pop_pmpm_pred <- read_csv(here::here("etc/outputs/cna_population_pred_costs.csv"))

# Read in demographic info
demographics <- read_csv(here::here("etc/outputs/demographics.csv")) 

# Demographic labels for more nuanced A2E plots later
demos_for_eval <- demographics %>% 
  mutate(meps_year = meps_year + 1) %>% 
  mutate(AGE_GRP = case_when(
    AGE_EXACT < 65 ~ "Under 65",
    AGE_EXACT < 70 ~ "65 - 69",
    AGE_EXACT < 75 ~ "70 - 74",
    AGE_EXACT < 80 ~ "75 - 79",
    AGE_EXACT < 85 ~ "80 - 84",
    AGE_EXACT >= 85 ~ "Over 85")) %>% 
  arrange(AGE_EXACT) %>% 
  mutate(AGE_GRP = forcats::fct_inorder(AGE_GRP)) %>% 
  select(-AGE_EXACT)

#### A-to-E plots for NE models - BY AGE ####
a2e_by_age_gender_NE.data <- cna_pop_pmpm_pred %>% 
  filter(model == "NE") %>% 
  left_join(demos_for_eval) %>% 
  group_by(AGE_GRP, SEX, version) %>% 
  summarize(Actual_pmpm = weighted.mean(cost_pmpm_act, w=exposure_months),
            Expected_pmpm = weighted.mean(cost_pmpm_pred, w=exposure_months),
            count = n()) %>%
  ungroup() %>% 
  pivot_longer(
    cols = -c(AGE_GRP, SEX, version, count),
    names_to = "Label",
    values_to = "Value",
    names_pattern = "^(Actual|Expected)_pmpm$"
  )  %>% 
  filter(!(Label == "Actual" & version == 28)) %>%  # v24 and v28 actuals are identical - need to remove one
  mutate(Label = if_else(Label == "Actual", "Actual", paste0("CMS-HCC v", version, " NE Predicted"))) %>% 
  mutate(Label = fct_inorder(Label)) %>% 
  mutate(AGE_GRP = fct_inorder(paste0(AGE_GRP, "\n(n = ", scales::label_comma(accuracy=1)(count), ")"))) %>% 
  mutate(Gender = if_else(SEX == "M", "Male", "Female"))

a2e_by_age_gender_NE.plot <- a2e_by_age_gender_NE.data %>% 
  ggplot(aes(x = AGE_GRP,
             y = Value,
             fill = Label)) + 
  geom_bar(stat="identity", position=position_dodge2(width=.9),
           color="gray20") + 
  geom_text(aes(label = scales::label_dollar(accuracy=1)(Value)),
            size = 3, position=position_dodge2(width=.9), vjust=-.8) +
  scale_fill_manual(values = cp) +
  scale_y_continuous(labels = scales::label_dollar(accuracy=1),
                     limits=c(0,1150)) +
  theme_bw() + 
  labs(
    title = "Actual Costs vs. Costs Predicted by CMS-HCC New Enrollee Models",
    subtitle = "Using Medical Expenditure Panel Data (MEPS) PUFs Datasets as Inputs",
    x = "Age Group",
    y = "Per Member, Per Month (PMPM) Costs (Medicare Only)",
    fill = "Value") +
  theme(legend.position = "bottom") +
  facet_wrap(Gender~., nrow=2, scales = "free_x")

a2e_by_age_gender_NE.plot

```

### New Enrollee Model - Predictive Ratios for Entire Population, by Race & Ethnicity

This isn't terrible performance, given that this is based on survey data. But risk adjuster models should be able to deliver accurate predictions for all "slices and dices" of the population. The accuracy of predictions breaks down rather quickly when we cut the data along groups that are irrelevant to the model (unlike as in the visualization in the previous section, where the score model's primary inputs are age and gender).

Since we're fortunate to be using MEPS data, we can cut the data on race and ethnicity for an interesting model performance evaluation.

```{r a2e_ne_race, echo=FALSE, fig.height=8.5, fig.width=11, message=FALSE, warning=FALSE}

# Read in cost projections 
cna_pop_pmpm_pred <- read_csv(here::here("etc/outputs/cna_population_pred_costs.csv"))

# Read in categorical factors and levels and join
survey_data_factors <- read_csv(here::here("etc/outputs/survey_data_factors.csv"))

#### A-to-E plots for NE models - BY AGE ####
a2e_by_raceth_NE.data <- cna_pop_pmpm_pred %>% 
  filter(model == "NE") %>% 
  left_join(survey_data_factors) %>% 
  group_by(version, RACETHX_DSC) %>% 
  summarize(Actual_pmpm = weighted.mean(cost_pmpm_act, w=exposure_months),
            Expected_pmpm = weighted.mean(cost_pmpm_pred, w=exposure_months),
            count = n()) %>%
  ungroup() %>% 
  pivot_longer(
    cols = -c(RACETHX_DSC, version, count),
    names_to = "Label",
    values_to = "Value",
    names_pattern = "^(Actual|Expected)_pmpm$"
  )  %>% 
  filter(!(Label == "Actual" & version == 28)) %>%  # v24 and v28 actuals are identical - need to remove one
  mutate(Label = if_else(Label == "Actual", "Actual", paste0("CMS-HCC v", version, " NE Predicted"))) %>% 
  arrange(RACETHX_DSC, Label) %>% 
  mutate(Label = fct_inorder(Label)) %>% 
  mutate(RACETHX_DSC= fct_inorder(paste0(str_wrap(RACETHX_DSC, 20), "\n(n = ", scales::label_comma(accuracy=1)(count), ")")))
                                  
                                  
a2e_by_raceth_NE.plot <- a2e_by_raceth_NE.data %>% 
  ggplot(aes(x = RACETHX_DSC,
             y = Value,
             fill = Label)) + 
  geom_bar(stat="identity", position=position_dodge2(width=.9),
           color="gray20") + 
  geom_text(aes(label = scales::label_dollar(accuracy=1)(Value)),
            size = 3, position=position_dodge2(width=.9), vjust=-.8) +
  scale_fill_manual(values = cp) +
  scale_y_continuous(labels = scales::label_dollar(accuracy=1)) +
  theme_bw() + 
  labs(
    title = "Actual Costs vs. Costs Predicted by CMS-HCC New Enrollee Models",
    subtitle = "Using Medical Expenditure Panel Data (MEPS) PUFs Datasets as Inputs",
    x = "Race & Ethnicity",
    y = "Per Member, Per Month (PMPM) Costs (Medicare Only)",
    fill = "Value") +
  theme(legend.position = "bottom")

a2e_by_raceth_NE.plot

```

The predictive accuracy here leaves much to be desired. Being able to predict member spending among diverse populations is an important primary objective of risk adjustment. For predictive modeling applications in healthcare, the NE just isn't very useful outside of predicting expenditures for populations where diagnosis information is not available. Some information, after all, is better than none at all.

Let's see if using a risk score model with a higher R-squared value will get our predictions closer to the pin...

## Predictive Performance of the Community Non-Dual Aged Model

The CMS-HCC Community Non-Dual Aged model is intended for use on Medicare beneficiaries who have been enrolled in Medicare for at least 12 months, over which time diagnosis code information can be gathered and used as inputs into the score model in order to increase predictive accuracy. 

### Dealing with truncated diagnosis codes in MEPS

Fortunately, MEPS data has ICD-10-CM diagnosis code information - but all diagnosis code info is truncated after the first three characters. This is extremely problematic for the project of risk scoring this population. For example, suppose a respondent in MEPS data is assigned the truncated code of ``E11``. In the v24 CMS-HCC model, this truncated diagnosis code could map to ``HCC 19 - Diabetes without complication`` which adds a scant .105 to the raw risk score. Or, the diagnosis code could map to ``HCC 18 - Diabetes with chronic complications`` which adds .302 to the raw risk score. Alternatively, the diagnosis code could map to ``HCC 17 - Diabetes with acute complications`` which adds a whopping 4.209 to the raw risk score. Without extra characters, we don't know.

To get around this limitation, using [publicly-available diagnosis code frequency tables from the state of California](https://data.chhs.ca.gov/organization/department-of-health-care-access-and-information), I built a probability table to "complete" each diagnosis code profile for each respondent in MEPS. (The sampling probabilities used are built from a frequency table which is constructed by the R script ``icd10-freq-prep.R`` in the ``src`` directory of this repo.)

Using the frequency distribution for each ICD-10-CM code, for each respondent, I simulate 500 possible combinations of all ICD-10-CM outcomes based on the truncated ICD-10-CM codes available in MEPS and tabulate the probability distribution of each possible diagnosis array. Then, I run each diagnosis array through the CMS-HCC risk score model, and take the probability-weighted average as the output for that respondent.

### Example of a simulated diagnosis profile from MEPS data

Consider respondent ID ``62197101`` who is in the MEPS datasets for the years 2019, 2020, and 2021. In 2019, the respondent has the following truncated ICD-10-CM codes listed in the medical conditions file: ``F32``, ``I10``, ``J40``, ``M54``, and ``R05``.

Using the frequency table, 500 probability-weighted combinations of fully completed diagnosis codes are sampled. The results of the sampling process are listed in the table below.

```{r dx_contingency, echo=FALSE, fig.height=8.5, fig.width=11, message=FALSE, warning=FALSE}

library(kableExtra)

# Load contingency dataset
load(here::here(paste0("etc/contingencies_n=500.rda")))

example_profile <- contingencies_to_run %>% 
  filter(DUPERSID == 62197101,
         meps_year == 2019) %>% 
  kbl() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

example_profile

```

At this point, all eight unique diagnosis profiles for this individual are risk scored by the CNA model *as if they were each separate people*. Afterwards, the score attributed to this respondent is the probability-weighted average of the risk score outputs for each of these eight diagnosis profiles.

### Community Non-Dual Aged Model - R-Squared for Entire Population, by MEPS Year

Similarly to how we computed R-squared values for model runs using the NE models by taking the square of correlation between raw score output and actual cost, we'll apply the same approach for the CNA models. Due to the use of diagnosis information, we expect a much larger R-Squared - though perhaps not the ~12% figure reported in the most recent [report to Congress](https://www.cms.gov/files/document/report-congress-risk-adjustment-medicare-advantage-december-2021.pdf), Table 2-1, due to the way we're limited from entirely accurate diagnosis information.

Note that there are two sets of model outputs for each version of the model. The *mode* score is the score that corresponds with the most frequently-occuring diagnosis profile associated with a respondent. The *average* score is the probability-weighted average of all scores resulting from running each diagnosis profile associated with a respondent. In the vast majority of cases, the mode score is equal to the average score, but this is, of course, not always the case.

```{r rsq_cna, echo=FALSE, fig.height=8.5, fig.width=11, message=FALSE, warning=FALSE}

# Read in cost projections 
cna_pop_pmpm_pred <- read_csv(here::here("etc/outputs/cna_population_pred_costs.csv"))

#### R-Squared Plot for NE Models by MEPS Year ####
# Evaluate r-sq by meps_year for NE. Should focus only on the pop. for which the NE is intended.
rsq_eval_CNA.data <- cna_pop_pmpm_pred %>% 
  filter(model == "CNA") %>% 
  group_by(meps_year, version, calculation) %>% 
  summarize(rsq = cor(cost_pmpm_act, cost_pmpm_pred)^2,
            count = n()) %>% 
  ungroup() %>% 
  mutate(meps_year = meps_year - 1) #Since this is an evaluation of scores, more intuitive to frame in terms of base year
  
rsq_eval_CNA.plot <- rsq_eval_CNA.data %>% 
  mutate(meps_year = fct_inorder(paste0(meps_year, "\n(n = ", scales::label_comma(accuracy=1)(count), ")"))) %>% 
  mutate(model = paste0("CMS-HCC v", version, " NE (", calculation, ")")) %>% 
  ggplot(aes(x = meps_year,
             y = rsq,
             fill = model)) +
  geom_bar(stat = "identity", position="dodge2",
           color="gray20") + 
  geom_text(aes(label=scales::label_percent(accuracy=.1)(rsq)),
            position=position_dodge2(width=.9), vjust=-.8, size=2.5) +
  theme_classic() +
  scale_fill_manual(values = cp) +
  scale_y_continuous(labels = scales::label_percent(accuracy=.1)) +
  labs(title = "Predictive Performance of CMS-HCC Community Non-Dual Aged Models",
       subtitle = "Using Medical Expenditure Panel Data (MEPS) PUFs Datasets as Inputs",
       y = "R-Squared",
       x = "MEPS Data Input Year",
       fill = "Model")

rsq_eval_CNA.plot
```

As we can see, model predictive performance (measured by R-Squared) improves substantially when using a model that builds risk score weights based on not just demographic factors, but also diagnosis code inputs.

### Community non-dual aged model - Predictive Ratios for Entire Population, by Age and Gender

As stated earlier, one of the most important uses of a risk adjuster model is to accurately predict costs. Let's see if cost projections broken out by age and gender groups using the more performant CNA model are closer to actuals than the predictions we got from the NE model earlier. Once again, we'll plot predicted vs. actual costs, grouped by age and gender, in order to evaluate how close (or far off) the predicted PMPM costs for each age band are, across all years evaluated. 

Our hope is that the CNA model predictions will come a lot closer to the actual outcomes than the NE model predictions...

```{r a2e_cna_age_gender, echo=FALSE, fig.height=8.5, fig.width=11, message=FALSE, warning=FALSE}

# Read in cost projections 
cna_pop_pmpm_pred <- read_csv(here::here("etc/outputs/cna_population_pred_costs.csv"))

# Read in demographic info
demographics <- read_csv(here::here("etc/outputs/demographics.csv")) 

# Demographic labels for more nuanced A2E plots later
demos_for_eval <- demographics %>% 
  mutate(meps_year = meps_year + 1) %>% 
  mutate(AGE_GRP = case_when(
    AGE_EXACT < 65 ~ "Under 65",
    AGE_EXACT < 70 ~ "65 - 69",
    AGE_EXACT < 75 ~ "70 - 74",
    AGE_EXACT < 80 ~ "75 - 79",
    AGE_EXACT < 85 ~ "80 - 84",
    AGE_EXACT >= 85 ~ "Over 85")) %>% 
  arrange(AGE_EXACT) %>% 
  mutate(AGE_GRP = forcats::fct_inorder(AGE_GRP)) %>% 
  select(-AGE_EXACT)

#### A-to-E plots for NE and CNA models - BY AGE/GENDER ####
a2e_by_age_gender_CNA.data <- cna_pop_pmpm_pred %>% 
  left_join(demos_for_eval) %>% 
  group_by(AGE_GRP, SEX, model, version, calculation) %>% 
  summarize(Actual_pmpm = weighted.mean(cost_pmpm_act, w=exposure_months),
            Expected_pmpm = weighted.mean(cost_pmpm_pred, w=exposure_months),
            count = n()) %>%
  ungroup() %>% 
  pivot_longer(
    cols = -c(AGE_GRP, SEX, model, version, calculation, count),
    names_to = "Label",
    values_to = "Value",
    names_pattern = "^(Actual|Expected)_pmpm$"
  ) %>% 
  filter((Label == "Actual" & version == 24 & calculation == "avg" & model == "CNA") | Label == "Expected") %>%  
  mutate(Label = if_else(Label == "Actual", "Actual", paste0("CMS-HCC v", version, " (", model, ", ", str_to_title(calculation), ")"))) %>% 
  mutate(Label = fct_inorder(Label)) %>% 
  mutate(AGE_GRP = fct_inorder(paste0(AGE_GRP, "\n(n = ", scales::label_comma(accuracy=1)(count), ")"))) %>% 
  mutate(Gender = if_else(SEX == "M", "Male", "Female"))

a2e_by_age_gender_CNA.plot <- a2e_by_age_gender_CNA.data %>% 
  ggplot(aes(x = AGE_GRP,
             y = Value,
             fill = Label)) + 
  geom_bar(stat="identity", position=position_dodge2(width=.9),
           color="gray20") + 
  geom_text(aes(label = scales::label_dollar(accuracy=1)(Value)),
            size = 3, position=position_dodge2(width=.9), vjust=-.8) +
  scale_fill_manual(values = cp) +
  scale_y_continuous(labels = scales::label_dollar(accuracy=1),
                     limits=c(0,1150)) +
  theme_bw() + 
  labs(
    title = "Actual Costs vs. Costs Predicted by CMS-HCC NE and CNA Models",
    subtitle = "Using Medical Expenditure Panel Data (MEPS) PUFs Datasets as Inputs",
    x = "Age Group",
    y = "Per Member, Per Month (PMPM) Costs (Medicare Only)",
    fill = "Value") +
  theme(legend.position = "bottom") +
  facet_wrap(Gender~., nrow=2, scales = "free_x")

a2e_by_age_gender_CNA.plot

```

...and that hope is confirmed by this plot, demonstrating that in almost every single age/gender category, the CMS-HCC CNA models outperform the NE models in producing accurate total cost projections!

### Community non-dual aged model - Predictive Ratios for Entire Population, by Race & Ethnicity

```{r a2e_cna_race, echo=FALSE, fig.height=8.5, fig.width=11, message=FALSE, warning=FALSE}

# Read in cost projections 
cna_pop_pmpm_pred <- read_csv(here::here("etc/outputs/cna_population_pred_costs.csv"))

# Read in categorical factors and levels and join
survey_data_factors <- read_csv(here::here("etc/outputs/survey_data_factors.csv"))

#### A-to-E plots for NE models - BY AGE ####
a2e_by_raceth_CNA.data <- cna_pop_pmpm_pred %>% 
  filter(model == "CNA" & calculation == "avg") %>% 
  left_join(survey_data_factors) %>% 
  group_by(version, RACETHX_DSC) %>% 
  summarize(Actual_pmpm = weighted.mean(cost_pmpm_act, w=exposure_months),
            Expected_pmpm = weighted.mean(cost_pmpm_pred, w=exposure_months),
            count = n()) %>%
  ungroup() %>% 
  pivot_longer(
    cols = -c(RACETHX_DSC, version, count),
    names_to = "Label",
    values_to = "Value",
    names_pattern = "^(Actual|Expected)_pmpm$"
  )  %>% 
  filter(!(Label == "Actual" & version == 28)) %>%  # v24 and v28 actuals are identical - need to remove one
  mutate(Label = if_else(Label == "Actual", "Actual", paste0("CMS-HCC v", version, " CNA Predicted"))) %>% 
  arrange(RACETHX_DSC, Label) %>% 
  mutate(Label = fct_inorder(Label)) %>% 
  mutate(RACETHX_DSC= fct_inorder(paste0(str_wrap(RACETHX_DSC, 20), "\n(n = ", scales::label_comma(accuracy=1)(count), ")")))
                                  
                                  
a2e_by_raceth_CNA.plot <- a2e_by_raceth_CNA.data %>% 
  ggplot(aes(x = RACETHX_DSC,
             y = Value,
             fill = Label)) + 
  geom_bar(stat="identity", position=position_dodge2(width=.9),
           color="gray20") + 
  geom_text(aes(label = scales::label_dollar(accuracy=1)(Value)),
            size = 3, position=position_dodge2(width=.9), vjust=-.8) +
  scale_fill_manual(values = cp) +
  scale_y_continuous(labels = scales::label_dollar(accuracy=1)) +
  theme_bw() + 
  labs(
    title = "Actual Costs vs. Costs Predicted by CMS-HCC Community Non-Dual Aged Models",
    subtitle = "Using Medical Expenditure Panel Data (MEPS) PUFs Datasets as Inputs",
    x = "Race & Ethnicity",
    y = "Per Member, Per Month (PMPM) Costs (Medicare Only)",
    fill = "Value") +
  theme(legend.position = "bottom")

a2e_by_raceth_CNA.plot

```

The CNA predictions by race and ethnicity are substantially improved over the predictions produced by the CMS-HCC NE model for the same cuts, for both versions 24 and 28. 

